%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This file is part of the LaTeX sources of the OMDoc 1.3 specification
% Copyright (c) 2006 Michael Kohlhase
% This work is licensed by the Creative Commons Share-Alike license
% see http://creativecommons.org/licenses/by-sa/2.5/ for details
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter*{Acknowledgments}
 \markboth{Acknowledgments}{Acknowledgments}\thispagestyle{empty}
Of course the {\omdoc} format has not been developed by one person
alone. The original proposal was taken up by several research
groups, most notably the {\OMEGA} group at Saarland University, the
{\maya} and {\activemath} projects at the German Research Center of
Artificial Intelligence (DFKI), the {\sc{MoWGLI}} EU Project, the
RIACA group at the Technical University of Eindhoven, and the
{\sc{CourseCapsules}} project at Carnegie Mellon University.  They
discussed the initial proposals, represented their materials in
{\omdoc} and in the process refined the format with numerous
suggestions and discussions.

The author specifically would like to thank Serge Autexier, Bernd Krieg-Br\"uckner, Olga
Caprotti, David Carlisle, Claudio Sacerdoti Coen, Arjeh Cohen, Armin Fiedler, Andreas
Franke, George Goguadze, Alberto Gonz\'alez Palomo, Dieter Hutter, Andrea Kohlhase,
Christoph Lange, Paul Libbrecht, Erica Melis, Till Mossakowski, Normen M\"uller, Immanuel
Normann, Martijn Oostdijk, Martin Pollet, Julian Richardson, Manfred Riem, and Michel
Vollebregt for their input, discussions, and feedback from implementations and
applications.

Special thanks are due to Alan Bundy and J\"org Siekmann. The first
triggered the work on {\omdoc}, has lent valuable insight over the
years, and has graciously consented to write the foreword to this
book. J\"org continually supported the {\omdoc} idea with his
abundant and unwavering enthusiasm. In fact the very aim of the
{\omdoc} format: openness, cooperation, and philosophic adequateness
came from the spirit in his {\OMEGA} group, which the author has had
the privilege to belong to for more than 10 years.

The work presented in this {\report} was supported by the ``Deutsche
For\-schungs\-gemeinschaft'' in the special research action ``Resource-adaptive cognitive
processes'' (SFB 378), and a three-year Heisenberg Stipend to the author.  Carnegie Mellon
University, SRI International, and Jacobs University Bremen have supported the author
while working on revisions for versions 1.1 to 1.3.


%%% Local Variables: 
%%% mode: stex
%%% TeX-master: "omdoc"
%%% End: 

% LocalWords:  MoWGLI RIACA CourseCapsules Autexier Bernd Krieg uckner Caprotti
% LocalWords:  Carlisle Sacerdoti Coen Arjeh Fiedler Franke Goguadze Gonz alez
% LocalWords:  Palomo Hutter Lange Libbrecht Melis Mossakowski Normen uller SFB
% LocalWords:  Normann Martijn Oostdijk Pollet Riem Vollebregt Bundy Joerg SRI
% LocalWords:  Siekmann Deutsche schungs gemeinschaft Br stex kohlhase markboth
% LocalWords:  thispagestyle omdoc activemath sc
