%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This file is part of the LaTeX sources of the OMDoc 1.3 specification
% Copyright (c) 2016 Michael Kohlhase.
% Source at https://github.com/KWARC/OMDoc/tree/master/doc/spec
% This work is licensed by the Creative Commons Share-Alike license
% see http://creativecommons.org/licenses/by-sa/2.5/ for details
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{oldpart}{MK: this chapter needs to be reworked totally to account for the document
    OMoC docuemnt models}
\begin{tchapter}[id=document-model]{Document Models for OMDoc}

  In almost all {\xml} applications, there is a tension between the document view and the
  object view of data; after all, {\xml} is a document-oriented interoperability framework
  for exchanging data objects. The question, which view is the correct one for {\xml} in
  general is hotly debated among {\xml} theorists. In {\omdoc}, actually both views make
  sense in various ways. Mathematical documents are the objects we try to formalize, they
  contain knowledge about mathematical objects that are encoded as formulae, and we arrive
  at content markup for mathematical documents by treating knowledge fragments (statements
  and theories) as objects in their own right that can be inspected and reasoned about.

  In {\mychaplref{mobj}{quiz}}, we have defined what {\omdoc} documents look like and
  motivated this by the mathematical objects they encode. But we have not really defined
  the properties of these documents as objects themselves (we will speak of the {\omdoc}
  {\twindef{document}{object model}} ({\indextoo{OMDOM}})). To get a feeling for the
  issues involved, let us take stock of what we mean by the object view of data. In
  mathematics, when we define a class of mathematical objects (e.g.  vector spaces), we
  have to say which objects belong to this class, and when they are to be considered equal
  (e.g.  vector spaces are equal, iff they are isomorphic). When defining the intended
  behavior of operations, we need to care only about objects of this class, and we can
  only make use of properties that are invariant under object equality. In particular, we
  cannot use properties of a particular realization of a vector space that are not
  preserved under isomorphism. For document models, we do the same, only that the objects
  are documents.


\begin{tsection}[id=xml-DOM]{XML Document Models}

  {\xml} supports the task of defining a particular class of documents (e.g. the class of
  {\omdoc} documents) with formal grammars such as the {\twintoo{document
      type}{definition}} ({\indextoo{DTD}}) or an {\xml} schema\twin{XML}{schema}, that
  can be used for mechanical document validation.  Surprisingly, {\xml} leaves the task of
  specifying document equality to be clarified in the (informal) specifications, such as
  this {\omdoc} specification. As a consequence, current practice for {\xml}
  applications\twin{application}{XML} is quite varied. For instance, the {\openmath}
  standard (see~\cite{BusCapCar:2oms04} and {\mysecref{openmath}}) gives a mathematical
  object model for {\openmath} objects that is specified independently of the {\xml}
  encoding. Other {\xml} applications like e.g.  presentation
  {\mathml}~\cite{CarIon:MathML03} or {\xhtml}~\cite{W3C:xhtml2000} specify models in form
  of the intended screen presentation, while still others like the
  {\xslt}~\cite{Clark:xslt99} give the operational semantics.

  For a formal definition let $\cK$ be a set of documents. We take a
  {\twindef{document}{model}} to be a partial equivalence relation\footnote{A partial
    equivalence relation is a symmetric transitive relation.  We will use $[d]_\cX$ for
    the {\twindef{equivalence}{class}} of $d$, i.e.  $[d]_\cX\colon=\{e|d{\cX}e\}$} $\cX$
  on documents, such that $\{d|d{\cX}d\}=\cK$. In particular, a relation $\cX$ is an
  equivalence relation on $\cK$.  For a given document model $\cX$, let us say that two
  documents $d$ and $d'$ are {\defemph{$\cX$-equal}}, iff $d{\cX}d'$. We call a property
  $p$ $\cX$-{\defemph{invariant}}\index{invariant!under a document model $\cX$}, iff for
  all $d{\cX}d'$, $p$ holds on $d$ whenever $p$ holds on $d'$.

  A possible source of confusion is that documents can admit more than one document model
  (see~\cite{KohKoh:esmk05} for an exploration of possible document models for
  mathematics). Concretely, {\omdoc} documents admit the {\omdoc} document model that we
  will specify in section {\mysecref{omdom}} and also the following four {\xml} document
  models that can be restricted to {\omdoc} documents (as a relation).\footnote{Here we
    follow Eliotte Rusty Harold's classification of layers of {\xml} processing
    in~\cite{Harold:ex03}, where he distinguishes the binary, lexical, sequence,
    structure, and semantic layer, the latter being the document model of the {\xml}
    application}
\begin{description}
\item[The {\twintoo{binary}{document model}}] interprets files as sequences of
  bytes.  Two documents are equal, iff they are equal as byte sequence. This
  is the most concrete and fine-grained (and thus weakest) document model
  imaginable.
\item[The {\atwintoo{lexical}{document}{model}}] interprets binary files as sequences of
  {\indextoo{Unicode}} characters~\cite{Unicode:tuc03} using an encoding table.  Two files
  may be considered equal by this document model even though they differ as binary files,
  if they have different encodings that map the byte sequences to the same sequence of
  {\unicode} characters.
\item[The {\xml} syntax document model\atwin{XML}{syntax}{document model}] interprets
  {\unicode} Files as sequences consisting of an {\xml} declaration, a DOCTYPE
  declaration, tags, entity references, character references, CDATA sections, PCDATA
  comments, and processing instructions. At this level, for instance, whitespace
  characters between {\xml} tags are irrelevant, and {\xml} documents may be considered
  the same, if they are different as {\unicode} sequences.
\item[The {\xml} structure document model\atwin{XML}{structure}{document model}]
  interprets documents as {\xml} trees of elements, attributes, text nodes, processing
  instructions, and sometimes comments. In this document model the order of attribute
  declarations in {\xml} elements is immaterial, double and single quotes can be used
  interchangeably for strings, and {\xml} comments\twin{XML}{comment}
  (\snippetin{<\char33--}\ldots\snippetin{-->}) are ignored.
\end{description}
Each of these document models, is suitable for different applications, for instance the
lexical document model is the appropriate one for {\indextoo{Unicode}}-aware editors that
interpret the encoding string in the {\xml} declaration and present the appropriate glyphs
to the user, while the binary document model would be appropriate for a simple ASCII
editor. Since the last three document models are refinements of the {\xml} document model,
we will recap this in the next section and define the {\omdoc} document model in
{\mysecref{omdom}}.

To get a feeling for the issues involved, let us compare the {\omdoc} elements in
{\mylstlref{first}{third}} below. For instance, the serialization in {\mylstref{second}}
is {\xml}-equal to the one in {\mylstref{first}}, but not to the one in
{\mylstref{third}}.

\begin{lstlisting}[escapechar=\%,label=lst:first,
   index={definition,CMP,OMOBJ,OMA},caption={An {\omdoc} Definition}]
<definition xml:id="comm-def" for="comm">
  <CMP xml:lang="en"><xhtml:p>
    An operation <OMOBJ id="op"><OMV name="op"/></OMOBJ>
     is called commutative, iff 
    <OMOBJ id="comm1">
      <OMA><OMS cd="relation1" name="eq"/>
        <OMA><OMV name="op"/><OMV name="X"/><OMV name="Y"/></OMA>
        <OMA><OMV name="op"/><OMV name="Y"/><OMV name="X"/></OMA>
      </OMA>
    </OMOBJ> for all <OMOBJ id="x"><OMV name="X"/></OMOBJ> 
    and <OMOBJ id="y"><OMV name="Y"/></OMOBJ>.</xhtml:p>
  </CMP>
  <CMP xml:lang="de">
    <xhtml:p>Eine Operation <OMOBJ><OMR href="#op"/></OMOBJ> hei%\ss%t kommutativ, falls 
    <OMOBJ><OMR href="#comm1"/></OMOBJ> f%\"u%r alle 
    <OMOBJ><OMR href="#x"/></OMOBJ> und 
    <OMOBJ><OMR href="#y"/></OMOBJ>.</xhtml:p>
  </CMP>
</definition>
\end{lstlisting}

\begin{lstlisting}[escapechar=\%,label=lst:second,mathescape,
   index={definition,CMP,OMOBJ},
   caption={An {\xml}-equal serialization for {\mylstref{first}}}]
<definition for="comm" xml:id="comm-def" >
  $\ldots$
  <CMP xml:lang='de'> <!-- Note the unabbreviated empty element -->
   <xhtml:p>Eine Operation <OMOBJ><OMR href="#op"/></OMOBJ>  hei%\ss%t 
   kommutativ, falls <OMOBJ><OMR href='comm1'/></OMOBJ> f%\"u%r alle 
   <OMOBJ><OMR href="#x"/></OMOBJ> und 
   <OMOBJ><OMR href='y'/></OMOBJ>.</xhtml:p>
  </CMP>
</definition>
\end{lstlisting}
\end{tsection}

\begin{tsection}[id=omdom]{The OMDoc Document Model}

  The {\omdoc} document model extends the {\xml} structure document model in various ways.
  We will specify the equality relation in the table below, and discuss a few general
  issues here.

  The {\omdoc} document model is guided by the notion of content markup for mathematical
  documents. Thus, two document fragments will only be considered equal, if they have the
  same abstract structure. For instance, the order of {\element{CMP}} children of an
  {\element{omtext}} element is irrelevant, since they form a
  {\twintoo{multilingual}{group}} which form the base for multilingual text
  assembly. Other facets of the {\omdoc} document model are motivated by
  presentation-independence, for instance the distribution of whitespace is irrelevant
  even in text nodes, to allow formatting and reflow in the source code, which is not
  considered to change the information content of a text.

\begin{lstlisting}[escapechar=\%,label=lst:third,
   index={definition,CMP,OMOBJ,OMA},
   caption={An {\omdoc}-Equal Representation for {\mylstsref{first}{second}}}]
<definition xml:id="comm-def" for="comm">
  <CMP xml:lang="de"><xhtml:p>Eine Operation <OMOBJ><OMR href="#op"/></OMOBJ> 
    hei%\ss%t kommutativ, falls 
    <OMOBJ id="comm1">
      <OMA><OMS cd="relation1" name="eq"/>
        <OMA><OMV name="op"/><OMV name="X"/><OMV name="Y"/></OMA>
        <OMA><OMV name="op"/><OMV name="Y"/><OMV name="X"/></OMA>
      </OMA>
    </OMOBJ> f%\"u%r alle <OMOBJ><OMR href="#x"/></OMOBJ> und 
    <OMOBJ><OMR href="#y"/></OMOBJ>.</xhtml:p>
  </CMP>
  <CMP xml:lang="en">
    <xhtml:p>An operation <OMOBJ id="op"><OMV name="op"/></OMOBJ>
    is called commutative, iff <OMOBJ><OMR href="#comm1"/></OMOBJ>
    for all <OMOBJ id="x"><OMV name="X"/></OMOBJ> and 
    <OMOBJ id="y"><OMV name="Y"/></OMOBJ>.</xhtml:p>
  </CMP>
</definition>
\end{lstlisting}

Compared to other document models, this is a rather weak (but general) notion of
equality. Note in particular, that the {\omdoc} document model does {\emph{not}}
use mathematical equality here, which would make the formula $X+Y=Y+X$ (the
{\element[ns-elt=om]{OMOBJ}} with {\snippet{xml:id="comm1"}} in
{\mylstref{third}} instantiated with addition for {\snippetin{op}}) mathematically
equal to the trivial condition $X+Y=X+Y$, obtained by exchanging the right hand
side $Y+X$ of the equality by $X+Y$, which is mathematically equal (but not
{\omdoc}-equal).

Let us now specify (part of) the equality relation by the rules in the table in
{\myfigref{omdom}}.  We have discussed a machine-readable form of these equality
constraints in the {\xml} schema for {\omdoc} in~\cite{KohAng:tccmvc03}.

\begin{myfig}{omdom}{The {\omdoc} Document Model}\scriptsize
\begin{tabular}{|l|p{1.2truecm}|p{5.5truecm}|p{3cm}|}\hline
  \#& Rule & comment & elements \\\hline\hline
  1 & unordered
  & The order of children of this element is irrelevant (as far as permitted by
  the content model). For instance only the order of {\element{obligation}}
  elements in the {\element{axiom-inclusion}} element  is arbitrary, since the
  others must precede them in the content model. 
  & {\element{adt}} {\element{axiom-inclusion}} {\element{metadata}}
  {\element{symbol}} {\element{code}} {\element{private}} {\element{presentation}}
  {\element{omstyle}}\\\hline
  2 & multi-group 
  & The order between siblings elements does not matter, as long as the values of
  the key attributes differ. 
  & {\element{CMP}} {\element{FMP}} {\element{requation}} {\element[ns-elt=dc]{description}}
  {\element{sortdef}} {\element{data}}   {\element[ns-elt=dc]{title}} {\element{solution}} \\\hline 
  3 & DAG encoding 
  & Directed acyclic graphs\atwin{directed}{acyclic}{graph} built up using
  {\element[ns-elt=om]{OMR}} elements are equal, iff their {\twintoo{tree}{expansion}s}
  are equal.  
  & {\element[ns-elt=om]{OMR}} \omdoc   reference \\\hline    
  4 & Dataset 
  & If the content of the {\element[ns-elt=dc]{type}} element is {\snippetin{Dataset}}, then
  the order of the siblings of the parent {\element{metadata}} element is
  irrelevant. 
  & {\element[ns-elt=dc]{type}} \\\hline
\end{tabular}
\end{myfig}

The last rule in {\myfigref{omdom}} is probably the most interesting, as we have seen in
{\mychapref{omdoc-infrastructure}}, {\omdoc} documents have both formal and informal
aspects, they can contain {\emin{narrative}} as well as {\emin{narrative-structured}}
information.  The latter kind of document contains a formalization of a mathematical
theory, as a reference for automated theorem proving systems. There, logical dependencies
play a much greater role than the order of serialization in mathematical objects.  We call
such documents {\defin{content OMDoc}} and specify the value {\snippetin{Dataset}} in the
{\element[ns-elt=dc]{type}} element of the {\omdoc} metadata for such documents.  On the
other extreme we have human-oriented presentations of mathematical knowledge, e.g.  for
educational purposes, where didactic considerations determine the order of
presentation. We call such documents {\defin{narrative-structured}} and specify this by
the value {\snippetin{Text}} (also see the discussion in {\mysecref{dc-elements}})
\end{tsection}

\begin{tsection}[id=sub-languages]{OMDoc Sub-Languages}

  In the last chapters we have described the {\omdoc} modules. Together, they make up the
  {\omdoc} document format, a very rich format for marking up the content of a wide
  variety of mathematical documents. (see {\mypartref{primer}} for some worked
  examples). Of course not all documents need the full breadth of {\omdoc} functionality,
  and on the other hand, not all {\omdoc} applications (see {\mypartref{applications}} for
  examples) support the whole language.

  One of the advantages of a modular language design is that it becomes easy to address
  this situation by specifying sub-languages that only include part of the
  functionality. We will discuss plausible {\omdoc} sub-languages and their applications
  that can be obtained by dropping optional modules from {\omdoc}.
  {\myfigref{omdoc-sub-languages}} visualizes the sub-languages we will present in this
  chapter. The full language {\omdoc} is at the top, at the bottom is a minimal
  sub-language {\omdoc} Basic, which only contains the required modules (mathematical
  documents without them do not really make sense). The arrows signify language inclusion
  and are marked with the modules acquired in the extension.

\begin{myfig}{omdoc-sub-languages}{{\omdoc} Sub-Languages and Modules}
 \fbox{\begin{tikzpicture}
     \begin{scope}
       \tikzstyle{every node}=[draw]
       \node (basic) at (3,0) {{\sc Basic} ({\small{\MOBJmodule{spec}},
           {\DOCmodule{spec}}, {\DCmodule{spec}}, {\MTXTmodule{spec}}, {\RTmodule{spec}}})};
       \node (cd) at (3,1.5) {{\sc{Content Dictionaries}}};
       \node (mathweb) at (1,2.8) {{\sc{MathWeb}}};
       \node (edu) at (1,4.4) {{\sc{Education}}};
       \node (spec) at (5,3.5) {{\sc{Specification}}};
       \node (full) at (3,5.5) {{\omdoc}};
     \end{scope}
    \draw[->] (basic) -- node[left]{\footnotesize{\PRESmodule{spec}}, {\STmodule{spec}}}  (cd);
    \draw[->] (cd) -- node[left]{\footnotesize{\EXTmodule{spec}}} (mathweb);
    \draw[->] (mathweb) --  node[left] {\footnotesize{\QUIZmodule{spec}}} (edu);
    \draw[->] (edu) -- 
        node[left] {\footnotesize{\CTHmodule{spec}}, {\DGmodule{spec}},
                                   {\PFmodule{spec}}, {\ADTmodule{spec}}}
              (full); 
    \draw [->] (cd) -- 
        node[right] {\footnotesize{\CTHmodule{spec}}, {\DGmodule{spec}}, 
                                  {\PFmodule{spec}}, {\ADTmodule{spec}}}
               (spec);
    \draw[->] (spec) -- node[right] {\footnotesize{\EXTmodule{spec}}, {\QUIZmodule{spec}}} (full);
  \end{tikzpicture}}
\end{myfig}

The sub-language identifiers can be used as values of the {\attributeshort{modules}}
attribute on the {\element{omgroup}} and {\element{omdoc}} elements. Used there, they
abbreviate the list of modules these sub-languages contain.

\begin{tsubsection}[id=sub-languages:basic]{Basic {\omdoc}}
  
  Basic {\omdoc} is sufficient for very simple mathematical documents that do not
  introduce new symbols or concepts, or for early (and non-specific) stages in the
  migration process from legacy representations of mathematical material (see
  {\mysecref{top-level}}).  This {\omdoc} sub-language consists of five modules: we need
  module {\MOBJmodule{spec}} for mathematical objects and formulae, which are present in
  almost all mathematical documents. Module {\DOCmodule{spec}} provides the document
  infrastructure, and in particular, the root element\twin{document}{root}
  {\element{omdoc}}. We need {\DCmodule{spec}} for titles, descriptions, and
  administrative metadata, and module {\MTXTmodule{spec}} so we can state properties about
  the mathematical objects in {\element{omtext}} element. Finally, module
  {\RTmodule{spec}} allows to structured text below the {\element{omtext}} level.  This
  module is not strictly needed for basic {\omdoc}, but we have included it for
  convenience.
\end{tsubsection}

\begin{tsubsection}[id=sub-languages:cd]{{\omdoc} Content Dictionaries}
  
  Content Dictionaries are used to define the meaning of {\indextoo{symbol}s} in the
  {\openmath} standard~\cite{BusCapCar:2oms04}, they are the mathematical documents
  referred to in the {\attribute{cd}{OMS}} attribute of the {\element[ns-elt=om]{OMS}}
  element.  To express content dictionaries in {\omdoc}\twin{content dictionary}{OMDoc},
  we need to add the module {\STmodule{spec}} to Basic {\omdoc}.  It provides the
  possibility to specify the meaning of basic mathematical objects ({\indextoo{symbol}s})
  by {\indextoo{axiom}s} and {\indextoo{definition}s} together with the infrastructure for
  inheritance, and grouping, and allows to reference the symbols defined via their home
  theory (see the discussion in {\mysecref{theories}}).
  
  With this extension alone, {\omdoc} content dictionaries add support for
  {\twintoo{multilingual}{text}}, simple {\indextoo{inheritance}} for theories, and
  {\twintoo{document}{structure}} to the functionality of {\openmath} content
  dictionaries. Furthermore, {\omdoc} content dictionaries allow the conceptual separation
  of mathematical properties into {\indextoo{constitutive}} ones and
  {\twintoo{logically}{redundant}} ones. The latter of these are not strictly essential
  for content dictionaries, but enhance maintainability and readability, they are included
  in {\openmath} content dictionaries for documentation and explanation.

  The sub-language for {\omdoc} content dictionaries also allows the specification of
  notations for the introduced symbols (by module {\PRESmodule{spec}}). So the resulting
  documents can be used for referencing (as in {\openmath}) and as a resource for deriving
  presentation information for the symbols defined here.  To get a feeling for this
  sub-language, see the example in the {\omdoc} variant of the {\openmath} content
  dictionary {\snippetin{arith1}} in {\mychapref{omcds}}, which shows that the {\openmath}
  content dictionary format is (isomorphic to) a subset of the {\omdoc} format. In fact,
  the {\openmath}2 standard only presents the content dictionary format used here as one
  of many encodings and specifies abstract conditions on content dictionaries that the
  {\omdoc} encoding below also meets.  Thus {\omdoc} is a valid content dictionary
  encoding.
\end{tsubsection}

\begin{tsubsection}[id=sub-languages:spec]{Specification {\omdoc}}
 
  {\omdoc} content dictionaries are still a relatively lightweight format for the
  specification of meaning of mathematical symbols and objects. Large scale formal
  specification efforts, e.g. for program verification need more structure to be
  practical. Specification languages like {\casl} (Common Algebraic
  Specification\index{specification} Language~\cite{CoFI:2004:CASL-RM}) offer the
  necessary infrastructure, but have a syntax that is not integrated with web standards.

  The Specification {\omdoc} sub-language adds the modules {\ADTmodule{spec}} and
  {\CTHmodule{spec}} to the language of {\omdoc} content dictionaries. The resulting
  language is equivalent to the {\casl} standard,
  see~\cite{AutHut:tefsduc00,Hutter:mocsv00,MAH-06-a} for the necessary theory.

The structured definition schemata from module {\ADTmodule{spec}} allow to specify
{\twintoo{abstract}{data type}s}, sets of objects that are inductively defined from
constructor symbols. The {\twintoo{development}{graph}} structure built on the
{\twintoo{theory}{morphism}s} from module {\CTHmodule{spec}} allow to make inclusion
assertions about theories that structure fragments of mathematical developments and
support a management of change\twin{management}{change}.
\end{tsubsection}

\begin{tsubsection}[id=sub-languages:mathwebomdoc]{MathWeb {\omdoc}}

{\omdoc} can be used as a content-oriented basis for web publishing of
mathematics. Documents for the web often contain images, applets, code fragments,
and other data, together with mathematical statements and theories.

The {\omdoc} sub-language MathWeb {\omdoc}\index{MathWeb {\sc{OMDoc}}} extends
the language for {\omdoc} content dictionaries by the module {\EXTmodule{spec}},
which adds infrastructure for images, applets, code fragments, and other data.
\end{tsubsection}

\begin{tsubsection}[id=sub-languages:web]{Educational {\omdoc}}
  
  {\omdoc} is currently used as a content-oriented basis for various systems for
  mathematics education (see e.g. {\mychapref{courseware}} for an example and discussion).
  The {\omdoc} sub-language Educational {\omdoc}\index{Educational {\sc{OMDoc}}}
  extends MathWeb {\omdoc} by the module {\QUIZmodule{spec}}, which adds
  infrastructure for exercises and assessments.
\end{tsubsection}

\begin{tsubsection}[id=spec-intro:reusing]{Reusing {\omdoc} modules in other formats}
  
  Another application of the modular language design is to share modules with other {\xml}
  applications\twin{XML}{application}. For instance, formats like
  {\docbook}~\cite{WalMue:dtdg2008} or {\xhtml}~\cite{W3C:xhtml2000} could be extended
  with the {\omdoc} statement level.  Including modules {\MOBJmodule{spec}},
  {\DCmodule{spec}}, and (parts of) {\MTXTmodule{spec}}, but not {\RTmodule{spec}} and
  {\DOCmodule{spec}} would result in content formats that mix the document-level structure
  of these formats.  Another example is the combination of {\xmlrpc} envelopes and
  {\omdoc} documents used for interoperability in {\mychapref{rpc}}.
\end{tsubsection}
\end{tsection}
\end{tchapter}
\end{oldpart}
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "spec"
%%% End: 

% LocalWords:  cd xxx arith omcds mathwebomdoc rpc ns omdom omdom omdom omdom
% LocalWords:  mobj OMDOM DOM openmath omdom Eliotte Harold's unicode PCDATA en
% LocalWords:  escapechar lst comm def lang cd eq Eine xref hei kommutativ alle
% LocalWords:  und basicstyle xsd adt  requation sortdef OMR Dataset href elt
% LocalWords:  dc mathescape concl omdom DOCTYPE CDATA omdom CMP OMOBJ OMA OMV
% LocalWords:  de omtext omdom omdom metadata omstyle multi FMP omdom omdoc edu

% LocalWords:  MathWeb MathWeb MathWeb
